Structure Learning for $VAR(p)$ Models

### 2019-03-15 paper.tex
Work on paper -- ancestor expansion.

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{P(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node

@@

\begin{theorem}[Granger Causality Equivalences]
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and uniformly bounded spectral density.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\mathcal{H}^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\mathcal{H}_t}$.  Then, the following are equivalent:

@@

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  That
is, $x_k$ may be a common ancestor that when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality, where $x_i \pwgc x_j$ if
$x_i$ Granger-causes $x_j$ with respect only to $(x_i, x_j)$.

We establish first some notation.  Remaining in the context of a
collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.  We refer to the
set of parents of $x_i$ in $\gcg$ by $\pa{i}$, that is,
$j \in \pa{i} \iff x_j \gc x_i$.  The set of \textit{ancestors} of
$x_i$ in $\gcg$ will be denoted $\anc{i}$, that is, $j \in \anc{i}$ if
and only if there is a directed path in $\gcg$ from $j$ to $i$.  Even
though the definitions of Granger-causality naturally suggests that
each node in the graph should have a self-loop (unless the node is
simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  We note that $\gcg$ need not be a DAG.

The edges of the graph can be given a general notion of ``weight'' by
associating an edge $(j, i)$ with an LTI filter
$\{B_{ij}(\tau)\}_{\tau = 1}^\infty$ from the $\VAR$ representation of equation
\ref{eqn:ar_representation}.

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A

\newcommand{\linE}[2]{\hat{\E}[#1\ |\ #2]}  % Linear projection
\newcommand{\linEerr}[2]{\xi[#1\ |\ #2]}  % Error of linear projection
\newcommand{\pa}[1]{pa(#1)}  % Parents of a node
\newcommand{\anc}[1]{\mathcal{A}(#1)}  % Ancestors of a node
\newcommand{\ancn}[2]{\mathcal{A}_{#1}(#2)}  % nth ancestors of a node
\newcommand{\gpn}[2]{gp_{#1}(#2)}  # nth generation grandparents
\newcommand{\gcgpath}[4]{#1 \rightarrow #2_1 \rightarrow \cdots \rightarrow #2_{#3} \rightarrow #4}

@@

\begin{theorem}[Granger Causality Equivalences]
  \label{thm:gcg}
  Let $x(t)$ be a WSS process with absolutely summable covariance
  sequence, and uniformly bounded spectral density.  Denote by
  $\xi_{ij} \defeq \linEerr{x_i(t)}{\mathcal{H}^{-j}}$ and
  $\xi_i \defeq \linEerr{x_i(t)}{\mathcal{H}_t}$.  Then, the following are equivalent:

@@

\subsection{Pairwise Granger Causality}
Recall that Granger-causality in general must be understood with
respect to a particular universe of observations.  If $x_j \gc x_i$
with respect to $x_{-k}$, it may not hold with respect to $x$.  For
example, $x_k$ may be a common ancestor which when observed completely
explains the connection from $x_j$ to $x_i$.  In particular, we will
study \textit{pairwise} Granger-causality.

\begin{definition}[Pairwise Granger-causality]
  We will say that $x_j$ pairwise Granger-causes $x_i$ and write
  $x_j \pwgc x_i$ if $x_i$ Granger-causes $x_j$ with respect only to
  $(x_i, x_j)$.
\end{definition}

We first need to establish some graph theoretic notation and
terminology, collected formally in a definition for the reader's
convenient reference.

\begin{definition}[Causality graph, Parents, etc...]

Remaining in the context of a collection of $n$ WSS processes
$x(t) = \big(x_1(t), \ldots, x_n(t)\big)$, we define the Granger-causality
graph $\gcg$ to be the directed graph formed on $n$ vertices where an
edge $(j, i) \in \gcg$ if and only if $x_j \gc x_i$.

We refer to the set of \textit{parents} of $x_i$ in $\gcg$ by
$\pa{i}$, that is, $j \in \pa{i} \iff x_j \gc x_i$.

Even though the definitions of Granger-causality naturally suggests
that each node in the graph should have a self-loop (unless the node
is simply independent noise) we will not include nodes in their own
parent sets: $i \not\in \pa{i} \subseteq \anc{i}$.  However, we do not
immediately assume that $\gcg$ is a DAG, and so we will allow for $i$
to be it's own \textit{ancestor}.

The set of level $\ell$ \textit{grandparents} of node $i$, denoted
$\gpn{\ell}{i}$, is the set such that $j \in \gpn{\ell}{i}$ if and only if
there is a \textit{directed and cycle-free path} of length $\ell$ in
$\gcg$ from $j$ to $i$ \textit{or} if $j = i$ and there is a cycle of
length $n$ starting at $i$ and ending back at $i$, that is, $i$ can be
it's own grandparent.  Clearly, $\pa{i} = \gpn{1}{i}$.

Finally, the set of \textit{level $\ell$ ancestors} of
$i$: $\ancn{\ell}{i} = \bigcup_{\lambda \le \ell}\gpn{\lambda}{i}$ is the set such that
$j \in \ancn{\ell}{i}$ if and only if there is a directed and cycle-free
path of length $\ell$ \textit{or less} in $\gcg$ from $j$ to $i$
\textit{or} if $j = i$ and there is a directed cycle of length $\ell$ or
less from $i$ back to $i$ in $\gcg$.  The set of \textit{all
  ancestors} of $i$ (e.g. $\ancn{n}{i}$) is denoted simply $\anc{i}$.
\end{definition}

The edges of the Granger-causality graph $\gcg$ can be given a general
notion of ``weight'' by associating an edge $(j, i)$ with a
\textit{strictly causal} LTI filter
$\B_{ij}(z) = \sum_{\tau = 1}^{\infty} B_{ij}(\tau)z^{-\tau}$ from the
$\VAR$ representation of equation \ref{eqn:ar_representation}.  We
will represent the action (convolution) of this filter compactly as

\begin{equation}
  \sum_{\tau = 1}^\infty B_{ij}(\tau)x_j(t - \tau) \defeq \B_{ij}(z)x_j(t).
\end{equation}

% What if we take this as the definition of PWGC?  then, in a simply causal graph
% the "normal" definition becomes a theorem.  Or, if lem:pwgc_anc is true then
% maybe that is good enough?

From the $\VAR$ representation of $x(t)$ there is clearly a tight relationship between each node and it's parent nodes, indeed

\begin{equation}
  \label{eqn:parent_expanion}
  x_i(t) = v_i(t) + \B_{ii}(z)x_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t).
\end{equation}

% When is this guaranteed w.r.t. the full system?
Furthermore, if the filter $\B_{ii}$ is invertible, we can write

\begin{equation*}
  \begin{aligned}
    (1 - \B_{ii}(z))x_i(t) &= v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\\
    \implies x_i(t) &= (1 - \B_{ii}(z))^{-1}\big[v_i(t) + \sum_{k \in \pa{i}}\B_{ik}(z)x_k(t)\big]\\
    &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)
  \end{aligned}
\end{equation*}

In general, we will write $\wtB$ for arbitrary filters maintaining the
convention that $\wtB_{ii}$ is a filter that modifies the future of
$x_i$ from it's past and $\wtB_{ij}$ a filter modifying $x_i$ from the
past of $x_j$.  We will use the convention that the filter
$\widetilde{\B}$ will represent simply ``a filter'' and is not
necessarily the same as the filters serving as edge weights in $\gcg$
(indeed, $\wtB$ will often be $0$), and moreover, we will allow for
$\widetilde{\B}$ to potentially change from place to place and even
from line to line in the same sequence of calculations as it's exact
properties or specification are not important.

We continue to develop the relationship between $x_i$ and it's family tree.

\begin{proposition}
  \label{prop:parent_expanding}
  The component $x_i(t)$ of $x(t)$ can be represented as a linear
  combination of a linear filtering operation acting on the driving
  noise $v_i$ of $x_i$ as well as the parents of $x_i$:

  \begin{equation}
    \label{eqn:parent_expansion}
    x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t).
  \end{equation}

  Moreover, $x_i$ can be expanded in terms of it's ancestors and
  driving noise only:

  \begin{equation}
    x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \anc{i} \setminus\{i\}}\wtB_{ik}(z)v_k(t).
  \end{equation}
\end{proposition}
\begin{proof}
  The formula \ref{eqn:parent_expansion} was established in the
  preceding paragraph and serves as a starting point for the induction
  hypothesis:

  \begin{equation}
    \label{eqn:induction}
    x_i(t) = \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
  \end{equation}

  The base case follows by expanding \ref{eqn:parent_expansion}, but we skip the calculations since they are nearly identical to the following induction step:

  \begin{align*}
    x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)\\
           &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell - 1}{i}\setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)\big[\wtB_{kk}v_k(t)\\
    &\ \ \ \  + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t)\big]\\
    &\overset{(b)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \ancn{\ell}{i} \setminus \{i\}}\wtB_{ik}(z)v_k(t) + \sum_{k \in \gpn{\ell + 1}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t).
  \end{align*}

  Where set $(a)$ follows by expanding each $x_k$ with equation \ref{eqn:parent_expansion} and $(b)$ requires the inversion of $\big(1 - \sum_{k \in \gpn{\ell}{i}\setminus\{i\}}\wtB_{ik}(z)\wtB_{ki}(z)\big)$ as well as using the fact that $\ancn{\ell}{i} = \ancn{\ell - 1}{i}\cup\gpn{\ell}{i}$.  We here remind the reader of our earlier established convention that $\wtB$ simply represents the existence of a filter (possibly 0) and can change from place to place.

The induction terminates after $n$ steps since in a graph with $n$ nodes $\gpn{n + 1}{i} = \emptyset$

% Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

%   \begin{align*}
%     x_i(t) &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)x_k(t)\\
%            &= \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)\big[\wtB_{kk}v_{k}(t) + \wtB_{ki}(z)x_i(t) + \sum_{h \in \pa{k}\setminus \{i\}}\wtB_{kh}(z)x_h(t) \big]\\
%            &\overset{(a)}{=} \wtB_{ii}(z)v_i(t) + \sum_{k \in \pa{i}}\wtB_{ik}(z)v_{k}(t) + \sum_{k \in \gpn{2}{i} \setminus \{i\}}\wtB_{ik}(z)x_k(t)
%   \end{align*}

% Where step $(a)$ requires the inversion of $1 - \sum_{k \in \pa{i}}\wtB_{ik}(z)\wtB_{ki}(z)$.

\begin{proposition}
  \label{prop:ancestor_properties}
  In a general Granger-causality graph $\gcg$, if $j \pwgc i$ then at
  least one of the following must hold

  \begin{enumerate}
    \item{$j \in \anc{i}$}
    \item{$\anc{i} \cap \anc{j} \ne \emptyset$}
  \end{enumerate}
\end{proposition}
\begin{proof}
\end{proof}

It seems intuitive that a converse of \ref{prop:ancestor_properties}
would hold, e.g. $j \in \anc{i} \implies j \pwgc i$.  Unfortunately,
this is not the case in general, as different paths through $\gcg$ can
lead to cancellation.

\begin{example}
  Firstly, on $n = 4$ nodes, ``diamond'' shapes can lead to cancellation on paths of length 2:

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{cccc}
      0 & 0 & 0 & 0\\
      a & 0 & 0 & 0\\
      -a & 0 & 0 & 0\\
      0 & 1 & 1 & 0\\
    \end{array}
  \right] x(t - 1) + v(t),
\end{equation*}

with $\E v(t) = 0,\ \E v(t)v(t - \tau)^\T = \delta_\tau I$.

The graph associated to this system is not strongly causal since there are two paths between $1$ and $4$, namely: $1 \rightarrow 2 \rightarrow 4$ and $1 \rightarrow 3 \rightarrow 4$.  However, by directly calculating

\begin{align*}
  x_4(t) &= x_2(t - 1) + x_3(t - 1) + v_4(t)\\
         &= ax_1(t - 2) + av_2(t - 1) - ax_1(t - 2) -av_3(t - 1) + v_4(t)\\
         &= a(v_2(t - 1) - v_3(t - 1)) + v_4(t),
\end{align*}

we see that, since $v(t)$ is spatially and temporally uncorrelated, $1 \not\pwgc 4$.
\end{example}

\begin{example}
  A second example on $n = 3$ nodes is also worth examining, in this
  cancellation is a result of differing time lags:

\begin{equation*}
  x(t) =
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      -a & 0 & 0\\
      0 & 1 & 0\\
    \end{array}
  \right] x(t - 1) +
  \left[
    \begin{array}{ccc}
      0 & 0 & 0\\
      0 & 0 & 0\\
      a & 0 & 0\\
    \end{array}
  \right] x(t - 2) + v(t)
\end{equation*}

Then

\begin{align*}
  x_2(t) &= v_2(t) - ax_1(t - 1)\\
  x_3(t) &= v_3(t) + x_2(t - 1) + ax_1(t - 2)\\
  \implies x_3(t) &= v_2(t - 1) + v_3(t),
\end{align*}

and again the graph is not strongly causal and $1 \not\pwgc 3$.
\end{example}

The previous examples seem rather pathological since we have simply
constructed a case where different paths in the graph cancel exactly.
It may be possible to rid ourselves of these pathologies for instance
by defining a notion of ``robust'' pairwise causality where
``$j \pwgc _R i$'' if $j \pwgc i$ for any ``small'' perturbation of
the system matrix.  However, we do not believe such a modification of
our theory can have useful practical implications since the difference
between paths that ``nearly cancel'' rather than cancel exactly would
in practice require inordinate amounts of data to resolve.

Alternatively, we could restrict the coefficients of the system
matrix, e.g. by requiring that $B_{ij}(\tau) \ge 0$.  Instead, we will
focus on the defining feature of time series networks, the topology of
$\gcg$ and consider conditions thereon.

\begin{definition}[Strongly Causal]
  We will say that a Granger-causality graph $\gcg$ is \textit{strongly causal}
if $\gcg$ is a DAG and there is at most 1 directed path between any two nodes.
\end{definition}

\begin{proposition}
  \label{prop:sc_graph_common_anc}
  In a strongly causal graph if $j \in \anc{i}$ and $k \in \anc{i} \cap \anc{j}$ then the unique path from $k$ to $i$ contains $j$.
\end{proposition}
\begin{proof}
  Suppose that there is a path from $k$ to $i$ which does not contain $j$.  In this case, there are multiple paths from $k$ to $i$ (one of which \textit{does} go through $j$ since $j \in \anc{i}$) which contradicts the assumption of strong causality.
\end{proof}

% Is pwgc transitive?

In light of proposition \ref{prop:sc_graph_common_anc}, the following provides a converse to proposition \ref{prop:ancestor_properties}.

\begin{proposition}
  \label{lem:pwgc_anc}
  If $\gcg$ is strongly causal and each self-loop filter
  $1 - \B_{ii}(z)$ is invertible, then $j \in \anc{i} \implies j \pwgc i$.
\end{proposition}
\begin{proof}
  Since strongly causal graphs are acyclic, this allows us to
  recursively expand the parents of $x_i$ in $\gcg$ in the
  $\VAR(\infty)$ representation of $x(t)$ since we will eventually
  terminate at nodes $d$ such that $\pa{d} = \emptyset$.  To this end,
  examining $x_i$ in equation \ref{eqn:ar_representation}

  \begin{align*}
    x_i(t) &= v_i(t) + \sum_{\tau = 1}^\infty \big[ B_{ii}(\tau)x_i(t - \tau) + \sum_{k \in \pa{i}} B_{ik}(\tau)x_k(t - \tau)\big]\\
    \implies x_i(t) &= (1 - \B_{ii}(z))^{-1} \big[v_i(t) + \sum_{k \in \pa{i}} B_{ik}(\tau)x_k(t - \tau) \big]\\
    &= \sum_{\tau = 1}^\infty \big[C_{ii}(\tau)v_i(t - \tau) + \sum_{k \in \pa{i}} C_{ik}(\tau)x_k(t - \tau)\big].
  \end{align*}

  If $\pa{i} = \emptyset$ then we simply have
  $x_i(t) = \sum_{\tau = 1}^\infty C_{ii}(\tau)v_i(t)$ and
  $C_{ii} = A_{ii}$ from equation \ref{eqn:wold}.

  Applying this relation recursively we can write

  \begin{equation*}
    x_i(t) = \sum_{\tau = 1}^\infty\big[C_{ii}(\tau)v_i(t - \tau) + \sum_{k \in \anc{i}}C_{ik}(\tau)v_k(t - \tau) \big].
  \end{equation*}

That strong conditions need to be placed on the topology of $\gcg$ in
order for lemma \ref{lem:pwgc_anc} to be true can be seen through the
earlier examples.

\begin{lemma}
  If $\gcg$ is strongly causal, then any $\VAR(p)$ system on $\gcg$ is stable if and only if
  $B_{ii}(z)$ is stable for every $i = 1, \ldots, n$.
\end{lemma}
\begin{proof}
\end{proof}

% $B(1) = \left[ \begin{array}{cc} 0 & 1 \\ 1 & 0 \end{array} \right]$
